---
execute:
  eval: false
---

# Cox regression model

This file contains R code for the analyses in Chapter 2 of the book Dynamic Prediction in Clinical Survival Analysis (CRC Chapman & Hall) by Hans C. van Houwelingen and Hein Putter

R code written by Hein Putter (H.Putter@lumc.nl for comments/questions)
The dynpred package is available from CRAN

Consistency with the book has been checked with
- R version 2.14.0
- survival version 2.36-10
- dynpred version 0.1.1


## Figure 2.1: Nelson-Aalen estimates of Data Sets 6 and 2

### Data set 6

```{r}
#| label: fig2.1_data6

require(dynpred)
data(ALL)
year <- 365.25

# Define relapse-free survival as the earliest of relapse and death
# Time to relapse (rel) and time to death (srv) are in days
# RFS will be in years
ALL$rfs <- pmin(ALL$rel,ALL$srv)/365.25
ALL$rfs.s <- pmax(ALL$rel.s,ALL$srv.s)

c0 <- coxph(Surv(rfs,rfs.s) ~ 1, data=ALL, method="breslow")
sf0 <- survfit(c0)
sf0 <- data.frame(time=c(0,sf0$time),surv=c(1,sf0$surv))
sf0$Haz <- -log(sf0$surv)
sf0 <- sf0[sf0$time<12,]
plot(sf0$time,sf0$Haz,type="s",xlab="Time (years)",ylab="Cumulative hazard",lwd=2)
```

### Data set 2

```{r}
#| label: fig2.1_data2

data(wbc1)

c0 <- coxph(Surv(tyears, d) ~ 1, data = wbc1, method="breslow")
sf0 <- survfit(c0)
sf0 <- data.frame(time=c(0,sf0$time),surv=c(1,sf0$surv))
sf0$Haz <- -log(sf0$surv)
sf0 <- sf0[sf0$time<8,]
plot(sf0$time,sf0$Haz,type="s",xlab="Time (years)",ylab="Cumulative hazard",lwd=2)
```

## Figure 2.2: Effect of the hazard ratio on the death risk

```{r}
#| label: fig2.2

F1seq <- seq(0,1,by=0.005) # this will be 1-S1
F2seq <- F1seq # this will be 1-S2; for HR=1 they are the same
plot(F1seq,F2seq,type="l",lwd=2,xlab="F1",ylab="F2")
HRs <- c(1,1.2,1.5,2,3)
for (i in 1:length(HRs)) {
    HR <- HRs[i]
    F2seq <- 1 - (1-F1seq)^HR
    lines(F1seq,F2seq,lwd=2,lty=i)
}
legend("topleft",paste("HR =",HRs),lwd=2,lty=1:5,bty="n")
```

## Figure 2.3: Comparing the effect of HR=1.5 (dotted versus solid) for different types of hazards

```{r}
#| label: fig2.3

tt <- seq(0,10,by=0.05)
HR <- 1.5 # Hazard ratio

oldpar <- par(no.readonly=TRUE) # save old graphical parameters
# Two plots
layout(matrix(1:2, 1, 2),widths=c(10.25,9))
# On the left, Weibull with shape=0.5, scale=5 (according to the
# parametrization used in R); this is a decreasing hazard
par(mar= c(5, 4, 4, 0.1) + 0.1)
sig <- 2
# The cumulative hazard is obtained using -log(survival)
H0 <- -pweibull(tt, 1/sig, 5, lower = FALSE, log = TRUE)
H1 <- H0 * HR
# Survival is obtained using reverse relation exp(-cumulative hazard)
plot(tt,exp(-H0),ylim=c(0,1),type="l",lwd=2,xlab="Time",ylab="Survival")
lines(tt,exp(-H1),type="l",lwd=2,lty=2)
title(main="Decreasing hazard")
# On the right, Weibull with shape=2, scale=5 (according to the
# parametrization used in R); this is an increasing hazard
par(mar= c(5, 0.1, 4, 1) + 0.1)
sig <- 0.5
H0 <- -pweibull(tt, 1/sig, 5, lower = FALSE, log = TRUE)
H1 <- H0 * HR
plot(tt,exp(-H0),ylim=c(0,1),type="l",lwd=2,xlab="Time",ylab="Survival",axes=FALSE)
lines(tt,exp(-H1),type="l",lwd=2,lty=2)
axis(1)
box()
title(main="Increasing hazard")
par(oldpar) # reset graphical parameters
```

## Example: Breast cancer II

Read in data (and provide labels); both data ("bc.txt" in tab delimited txt format and syntax should be in working directory for this to work

```{r}
#| label: read_bc
source("bc syntax.R")
```

## Table 2.1: The Cox model for the EORTC breast cancer data (Data Set 5), including all risk factors

```{r}
#| label: table2.1

cfull <- coxph(Surv(survyrs,survstat) ~ surgery + tusi + nodal + age50 +
  adjchem + tam + periop, data=bc, method="breslow")
cfull

Xbeta <- bc$Xbeta <- cfull$linear.predictors
mean(bc$Xbeta) # is 0
sd(bc$Xbeta)
```

## Figure 2.4: Histogram of the prognostic index PI; in dark-grey patients at risk after eight years, in light-grey patients who died or were lost-to-follow-up within eight years

```{r}
#| label: fig2.4

cuts <- seq(-2,2,by=0.5)
bc$risk8 <- 1
bc$risk8[bc$survyrs>=8] <- 0
tbl <- table(bc$risk8,cut(bc$Xbeta,cuts))
barplot(tbl,space=0,ylim=c(0,1000),xlab="Prognostic index",ylab="Frequency",axes=FALSE,axisnames=FALSE)
axis(1,at=0:8,labels=cuts)
axis(2)
box()
```

## Distribution of prognostic index after 8 years

```{r}
#| label: progind

table(bc$risk8) # bc$risk8=0 means at risk
# Those still at risk
mean(bc$Xbeta[bc$risk8==0])
sd(bc$Xbeta[bc$risk8==0])
# Those no longer at risk
mean(bc$Xbeta[bc$risk8==1])
sd(bc$Xbeta[bc$risk8==1])
```


## Figure 2.5: Predicted survival curves for different values of the prognostic index


```{r}
#| label: fig2.5

# Rerun a Cox model with the prognostic index
# This should have an estimated coefficient of exactly 1 (check)
cXbeta <- coxph(Surv(survyrs,survstat) ~ Xbeta, data=bc, method="breslow")

# The survfit function from the survival library can take a "newdata"
# argument, in which the value(s) of the covariate(s) is/are specified
nd <- data.frame(Xbeta=-2*sd(Xbeta))
sf <- survfit(cXbeta,newdata=nd)
plot(sf,lwd=2,conf.int=FALSE,mark.time=FALSE,lty=2,xlab="Time in years",ylab="Survival function")
nd <- data.frame(Xbeta=-1*sd(Xbeta))
sf <- survfit(cXbeta,newdata=nd)
lines(sf,lwd=2,conf.int=FALSE,mark.time=FALSE,lty=3)
nd <- data.frame(Xbeta=0)
sf <- survfit(cXbeta,newdata=nd)
lines(sf,lwd=2,conf.int=FALSE,mark.time=FALSE,lty=1)
nd <- data.frame(Xbeta=sd(Xbeta))
sf <- survfit(cXbeta,newdata=nd)
lines(sf,lwd=2,conf.int=FALSE,mark.time=FALSE,lty=4)
nd <- data.frame(Xbeta=2*sd(Xbeta))
sf <- survfit(cXbeta,newdata=nd)
lines(sf,lwd=2,conf.int=FALSE,mark.time=FALSE,lty=5)

# Add Kaplan-Meier plot
c.km <- coxph(Surv(survyrs,survstat) ~ 1, data=bc, method="breslow")
sf <- survfit(c.km)
lines(sf,lwd=2,conf.int=FALSE,mark.time=FALSE,col=8)
legend("bottomleft",c("Mean-2sd","Mean-sd","Mean","Mean+sd","Mean+2sd","Kaplan-Meier"),lwd=2,col=c(1,1,1,1,1,8),lty=c(2,3,1,4,5,1),bty="n")
```

## Figure 2.6: Dynamic prediction curves for different values of the prognostic index

```{r}
#| label: fig2.6

nd <- data.frame(Xbeta=2*sd(Xbeta))
sf <- survfit(cXbeta,newdata=nd)
Fw <- Fwindow(sf,5)
# Plot first 8 years only
Fw <- Fw[Fw$time<=8,]
plot(Fw$time,Fw$Fw,type="s",ylim=c(0,0.5),xlab="Years surviving",ylab="Probability of dying within window",lwd=2,lty=5)
nd <- data.frame(Xbeta=sd(Xbeta))
sf <- survfit(cXbeta,newdata=nd)
Fw <- Fwindow(sf,5)
Fw <- Fw[Fw$time<=8,]
lines(Fw$time,Fw$Fw,type="s",lwd=2,lty=4)
nd <- data.frame(Xbeta=0)
sf <- survfit(cXbeta,newdata=nd)
Fw <- Fwindow(sf,5)
Fw <- Fw[Fw$time<=8,]
lines(Fw$time,Fw$Fw,type="s",lwd=2,lty=1)
nd <- data.frame(Xbeta=-sd(Xbeta))
sf <- survfit(cXbeta,newdata=nd)
Fw <- Fwindow(sf,5)
Fw <- Fw[Fw$time<=8,]
lines(Fw$time,Fw$Fw,type="s",lwd=2,lty=3)
nd <- data.frame(Xbeta=-2*sd(Xbeta))
sf <- survfit(cXbeta,newdata=nd)
Fw <- Fwindow(sf,5)
Fw <- Fw[Fw$time<=8,]
lines(Fw$time,Fw$Fw,type="s",lwd=2,lty=2)
c.km <- coxph(Surv(survyrs,survstat) ~ 1, data=bc, method="breslow")
sf <- survfit(c.km)
Fw <- Fwindow(sf,5)
Fw <- Fw[Fw$time<=8,]
lines(Fw$time,Fw$Fw,type="s",lwd=2,col=8)
legend("topright",c("Mean-2sd","Mean-sd","Mean","Mean+sd","Mean+2sd","Kaplan-Meier"),lwd=2,col=c(1,1,1,1,1,8),lty=c(2,3,1,4,5,1),bty="n")
```

## Delayed entry (code and data is availabe upon request)

